\section{Variance Bounds}
\label{app:variance}

This appendix expands the variance calculation used in Lemma~\ref{lem:message-recurrence} and Theorem~\ref{thm:main}. Suppose a scalar tensor contraction can be written as
\[
p=\sum_{\omega} q(\omega) z(\omega)
\]
where \(q\) is a signed or complex quasiprobability representation. Let
\[
\gamma=\sum_{\omega}|q(\omega)|.
\]
Sampling \(\omega\) from \(Q(\omega)=|q(\omega)|/\gamma\) and returning
\[
Z=\gamma\,\operatorname{phase}(q(\omega))\,z(\omega)
\]
gives \(\E[Z]=p\). If \(|z(\omega)|\le B\), then
\[
\E[|Z|^2]\le \gamma^2 B^2.
\]
Thus \(N=O(\gamma^2B^2\eps^{-2})\) samples suffice for constant-probability additive estimation, with logarithmic amplification for failure probability \(\delta\).

In global quasiprobability simulation, \(\gamma\) is the product of all local \(\ell_1\) costs. In SLMS, Lemma~\ref{lem:message-recurrence} proves for locally marginalizable networks that the effective \(\gamma\) for the hardest message is bounded by
\[
\gamma_{\mathrm{eff}}\le \exp(\Lammsg(C,\T)).
\]
Consequently the sampling term in Theorem~\ref{thm:main} is explicitly \(\exp(2\Lammsg(C,\T))\). The bounded term \(B\) is accounted for by the effective active boundary dimension \(2^{O(\weff)}\) and by normalization conventions for local tensors.

\subsection{Median-of-means}

Let \(Z_1,\ldots,Z_N\) be independent unbiased estimators with variance at most \(\sigma^2\). Partition the samples into \(K=O(\log(1/\delta))\) groups, each of size \(M=O(\sigma^2/\eps^2)\). Compute the mean in each group and return the median. Standard median-of-means analysis gives additive error \(\eps\) with failure probability at most \(\delta\). This avoids requiring bounded support beyond a second-moment bound.

\subsection{Message-table estimators}

The main theorem estimates the root scalar. If an implementation instead estimates intermediate messages as random tensors, it must choose a norm in which to control their second moments and must include the active-boundary table dimension \(D_b\le 2^{O(\weff)}\), per-entry accuracy, and union bounds. The local recurrence in Lemma~\ref{lem:message-recurrence} controls normalized entries or normalized linear functionals through the certified coefficient-norm factors \(\kappa_b\). Replacing these local certificates by a sharper general separator rule is part of the broader separator-factorization conjecture.
