\section{Worked Certificate on a Fallback Lambeq-Compatible Structured Instance}
\label{app:worked-experiment-certificate}

This appendix records one complete certificate produced by the artifact command recorded in \texttt{results/run\_manifest.json}.
The selected instance is \(\texttt{fallback\_0000}\). The real lambeq attempt produced 120 local LinearReader graph-only diagrams, but the available Bobcat/DepCCG/CCG parser paths did not produce parsed diagrams for the sample corpus and the reader graphs contain no non-Clifford resource annotations. This appendix is therefore deliberately titled as a fallback lambeq-compatible structured instance rather than a real parsed lambeq resource certificate.

\paragraph{Instance.}
The instance has identifier \(\texttt{fallback\_0000}\), mode \(\texttt{leaf\_local}\), and sentence label
\[
\texttt{fallback\_leaf\_local\_sentence\_with\_16\_word\_modules}.
\]
It is a balanced binary parse tree with \(31\) bags, \(16\) word leaves, and \(30\) decomposition edges. The root is \(b0\). The internal bags \(b0,\ldots,b14\) are free grammar-composition tensors. The leaf bags \(b15,\ldots,b30\) contain locally consumed resource word modules.

\paragraph{Graph and decomposition.}
The graph is the parse tree itself. The decomposition uses one bag per parse-tree node. The parent map is the standard heap-style binary-tree map:
\[
p(b_i)=b_{\lfloor (i-1)/2\rfloor}\qquad (1\le i\le 30).
\]
The resource assignment \(\alpha\) assigns one resource tensor to each leaf bag \(b15,\ldots,b30\) and no resource tensor to internal bags.

\paragraph{Local costs and active-child certificate.}
Each leaf resource has certified cost \(\xi_b=1.3\), each internal bag has \(\xi_b=1\), and every local coefficient map has \(\kappa_b=1\). Every leaf resource is locally measured or marginalized into a bounded free feature message before crossing its first separator. Hence all active-child sets are empty:
\[
\Act(b)=\emptyset\qquad\text{for every bag }b.
\]
The recurrence therefore gives \(\Gamma_b=1.3\) at resource leaves and \(\Gamma_b=1\) at internal bags. Consequently
\[
\Lammsg=\log(1.3)=0.262364\ldots,
\qquad
\Lambda_{\mathrm{glob}}=16\log(1.3)=4.197828\ldots .
\]
The heuristic dense treewidth proxy reported by the artifact is \(\wTN=1\), the effective-width proxy is \(\weff=1\), and the exponent proxies are
\[
\weff+2\Lammsg=1.524729\ldots,
\quad
2\Lambda_{\mathrm{glob}}=8.395656\ldots,
\quad
\weff+2\Lambda_{\mathrm{glob}}=9.395656\ldots .
\]
The focused-width proxy for this instance is \(1.0\), using the implemented resource-separator proxy rather than exact focused tree-width.

\begin{center}
\small
\begin{tabular}{lcccccc}
\toprule
Bag set & Resources & Count & \(\xi_b\) & \(\kappa_b\) & \(\Act(b)\) & \(\Gamma_b\) \\
\midrule
\(b0,\ldots,b14\) & no & 15 & 1.0 & 1.0 & \(\emptyset\) & 1.0 \\
\(b15,\ldots,b30\) & yes & 16 & 1.3 & 1.0 & \(\emptyset\) & 1.3 \\
\bottomrule
\end{tabular}
\end{center}

The artifact also writes the bag-by-bag recurrence table to
\[
\texttt{results/worked\_certificate\_recurrence.csv}.
\]
For this instance every row of that CSV is determined by the two grouped rows above. This appendix demonstrates the contents of an SLMS certificate: graph/decomposition data, resource assignment, finite local costs, local \(\kappa_b\) factors, active-child declarations, recurrence values, and baseline exponent proxies.
