\section{Reproducibility and Artifact Guide}
\label{app:reproducibility-artifact}

The release artifact is intended to accompany the paper as executable evidence for the certificate computations, not as evidence of general parser-level QNLP coverage. The public repository URL is \url{https://github.com/yazhsab/slms-separator-local-magic}, and the archival record is Zenodo DOI \href{https://doi.org/10.5281/zenodo.19805000}{10.5281/zenodo.19805000}.

\paragraph{One-command certificate run.}
From the repository root, run
\[
\texttt{python experiments/slms\_certificate/run\_all.py --mode auto}.
\]
The verified environment used a Python 3.11 virtual environment at \texttt{/tmp/slms\_lambeq\_py311}; the same command also works with a compatible Python environment containing the packages listed in \texttt{experiments/slms\_certificate/requirements.txt}.

\paragraph{Real lambeq mode and fallback mode.}
The \texttt{--mode real} path attempts parser-generated lambeq extraction and fails honestly if parser backends or model files are unavailable. The \texttt{--mode fallback} path runs fallback structured lambeq-compatible instances. The \texttt{--mode auto} path tries real extraction first and then falls back, recording the selected path in \texttt{results/run\_manifest.json}. In the executed run reported in this paper, parser-generated lambeq/QDisCoCirc resource instances were unavailable; fallback structured instances are therefore the primary nontrivial certificate evidence.

\paragraph{Expected outputs.}
The main output files are
\[
\begin{array}{ll}
\texttt{results/slms\_certificate\_results.csv} & 120\ \text{rows},\\
\texttt{results/slms\_certificate\_failures.csv} & 15\ \text{rows},\\
\texttt{results/baseline\_results.csv} & 1680\ \text{rows},\\
\texttt{results/fallback\_instances.json} & 120\ \text{instances},\\
\texttt{results/real\_lambeq\_instances.json} & 120\ \text{LinearReader sidecar instances},\\
\texttt{results/worked\_certificate\_recurrence.csv} & \text{worked certificate rows}.
\end{array}
\]
The plots are regenerated with
\[
\texttt{python experiments/slms\_certificate/plot\_results.py}
\]
and written to \texttt{figures/}.

\paragraph{Hard blockers.}
The files \texttt{hard\_blocker\_ledger.md}, \texttt{lambeq\_parser\_blocker\_report.md}, \texttt{zx\_baseline\_blocker\_report.md}, and \texttt{stn\_mast\_blocker\_report.md} record unavailable external components. In particular, the artifact does not contain parser-generated lambeq/QDisCoCirc resource instances, an exact Sutcliffe--Kissinger ZX cutting or partitioning implementation, or exact Masot-Llima/Nakhl STN/MAST implementations. The reported ZX and STN rows are proxy diagnostics unless explicitly stated otherwise.
